Parallelizing AdaBoost and Decision Trees

### 2015-09-04 paper.tex
paper additions

\indent Decision trees are a well-known method used in classification which partitions
the feature space into rectangular regions and gives each region the most
frequent label among the training set points it contains. In the most common
decision tree algorithms each partition of the feature space is made along
coordinate axes and done in a heirarchical way.  Starting with the entire
space, a line parallel to an axis divides it into two regions and then the same
is done for these resulting regions. This process is repeated many times until
the space is partitioned into several rectangles, like the example in Figure 1
(\cite{HTF}). Each split is made to separate points of the two distinct classes
into different regions as best as possible. Thus the two resulting rectangles
will have a greater {\it purity} in comparison to the parent node.

\begin{figure}[b!]
\centering
\begin{minipage}{0.45\textwidth}
\centering
\includegraphics[width=2.5in]{partition}
\caption{space partitioned by a decision tree}
\end{minipage}\hfill
\begin{minipage}{0.45\textwidth}
\centering
\includegraphics[width=2.5in]{dectree}
\caption{decision tree}
\end{minipage}
\end{figure}

There are a few common ways to measure purity for a given region containing
sample points. Gini Index is used here which can be easily substituted for
others. For two classes the Gini Index of region $A$ is $I(A) = 2p(1-p)$ where
$p$ is the probability of choosing a point in the first class from $A$, which
translates to the probability of incorrectly guessing the class of a randomly
chosen point from either class.

The splitting of regions gives rise to the tree structure. At each node in the
tree, the samples in the region corresponding to that node are ordered and
counted to find in which dimension that region can be divided that gives the
greatest increase in the purity of the branch. Figure 2 shows an example of
this where at the top of the tree the decision to split on the first feature
$X_1$ is best and the separation is made at $t_1$. In the first of the two
resulting regions it is determined that the best split is made on $X_2$ at
$t_2$, and in the other region the best separation is on $X_1$ at $t_3$. This
splitting process will go on until a stopping criterion is met, either if there
are too few sample points in a region or if no gain in purity can be made by
splitting on any feature.

@@

Since the tree-building process is the most time-consuming portion of AdaBoost,
our goal is to parallelize this. It seems natural to divide amongst processors
the work of finding the dimension in the feature space that best splits the
region and increases the purity for each node in the tree. The datasets we are
handling have several dimensions, on the order of tens to hundreds, so each
processor could be responsible for a group of dimensions in the feature space.
After each processor determines the best dimension to split amongst its
group, we can perform a reduction and determine the best split over all
dimensions.

@@

%\bibitem{no4} Wiki
%\end{thebibliography}
\end{document}

>>>>

\indent Decision trees are a well-known method used in classification which
partitions the feature space into rectangular regions and applies to each
region the most frequent label among the training set points it contains. In
the most common decision tree algorithms each partition of the feature space is
made along coordinate axes and is done in a heirarchical way. Initially, a line
parallel to an axis divides the space into two regions, and then the same is
done for these resulting regions. This process is repeated many times until the
space is partitioned into several rectangles, like the example in Figure 1
(\cite{HTF}). Each split is made to separate points of the two distinct classes
into different regions as best as possible. Thus the two resulting rectangles
will have less {\it impurity} in comparison to the parent node.

There are a few common ways to measure impurity for a given region containing
sample points. Gini index is used here which can be easily substituted for
others. For two classes the Gini Index of region $A$ is $I(A) = 2p(1-p)$ where
$p$ is the probability of choosing a point in the first class from $A$, which
translates to the probability of incorrectly guessing the class of a randomly
chosen point from either class.

The splitting of regions gives rise to the tree structure. At each node in the
tree, the samples in the region corresponding to that node are ordered and
counted to find in which dimension that region can be divided that gives the
greatest decrease in the impurity of the branch. Figure 2 shows an example of
this where at the top of the tree the decision to split on the first feature
$X_1$ is best and the separation is made at $t_1$. In the first of the two
resulting regions it is determined that the best split is made on $X_2$ at
$t_2$, and in the other region the best separation is on $X_1$ at $t_3$. This
splitting process will go on until a stopping criterion is met, either if there
are too few sample points in a region or if no gain in purity can be made by
splitting on any feature.

@@

\begin{figure}[h!]
\centering
\begin{minipage}{0.45\textwidth}
\centering
\includegraphics[width=2.5in]{partition}
\caption{space partitioned by a decision tree}
\end{minipage}\hfill
\begin{minipage}{0.45\textwidth}
\centering
\includegraphics[width=2.5in]{dectree}
\caption{decision tree structure}
\end{minipage}
\end{figure}

\section{Implementation}

Since the tree-building process is the most time-consuming portion of AdaBoost,
our goal is to parallelize this. It seems natural to divide amongst processors
the work of finding the dimension in the feature space that best splits the
region and increases the purity for each node in the tree. The datasets we are
handling have several dimensions, on the order of tens to hundreds, so each
processor could be responsible for a group of dimensions in the feature space.

The AdaBoost algorithm is written in the main method of the program contained
in {\tt paraboost.c}. This file also contains a few methods for utility, one
that converts a decision tree data structure into a useable function and
another that calculates weighted error. main() also performs load balancing by
determing the features that each processor is responsible for.

The methods that collect data from files on the hard disk and organize storage
are defined in {\tt data.c}. These methods scrape floats from text and binary
files and allocate space on the heap to store feature vectors and their labels.
main() calls them early to load the training data in memory. There are a number
of different sorting routines in {\tt sort.c}, all implementations of QuickSort
or MergeSort that vary with the structure of the data types they sort. These
methods are used to order data structures prior to the decision tree
construction.

The methods that implement decision tree operations are located in {\tt
tree.c}. ParallelSplit() performs the parallel node splitting procedure for a
given node in a decision tree, using the data points in that node to find the
partition minimizing the impurity of the resulting child nodes. It calls the
subroutine PodWBS() for each feature, which computes the impurity for every
possible partition of the data over that feature dimension and returns an index
corresponding to the best partition for that dimension and the impurity. Since
this is done for every feature the processor is responsible for,
ParallelSplit() then broadcasts which of its features is best and the minimum
impurity to every processor. Thus an Allreduce is performed in order to
determine which feature among all has the best partition achieving the minimum
impurity. This determines the two branches of the decision tree, and the root
processor informs the others which branch of the tree each observation goes to.
The arrays of pointers to these observations are re-sorted to reflect this and
ParallelSplit() is then called recursively on each of the two child nodes.
After an end condition is met (depth in tree, small number of points
in node, or very low impurity) the function returns. Thus this method is called for every node in the tree and performs noticeable computation for all nodes that aren't leaves.

%After each processor determines the best dimension to split on amongst those
%it's responsible for, we can perform a reduction and determine the best split
%over all dimensions.

Any tree will split the feature space multiple times so
there are several occasions for parallelism to speed up execution. In testing I
commonly set the maximum depth for a tree to between two to five levels. Also
the number of rounds of AdaBoost, and hence the number of trees constructed, is
a parameter in the method allowing you to affect how often the algorithm executes in parallel.

\section{Data and Results}
We employ the MNIST dataset to test our implementation. MNIST is a collection of 70,000 images of handwritten digits commonly used in image processing and pattern recognition research. The grayscale images are written in a binary file and consist of 28x28 arrays of integers ranging from 0 to 255 representing pixels varying from white to black. From this we separate out the set of 1's and 7's with the goal of performing binary classification between these two digits. We'll also trying to do the same thing with the digits 4 and 9.

With the MNIST dataset, a speedup by a factor of roughly four can be achieved
when using four processors, yet increasing beyond this does not give any
noticeable gains.

@@

%\bibitem{no4} Wiki AdaBoost

%\bibitem{MNIST} Yann LeCun MNIST

%\end{thebibliography}
\end{document}
